---
output: github_document
---

### NSGEV 

The **NSGEV** R package is devoted to some Non-Stationary 
Extreme Value models

The main function of the package is `TVGEV` which creates an object
with class `"TVGEV"` representing a Time-Varying model with GEV
margins depending on the time. This kind of model is especially useful
to study *block maxima*, usually annual maxima.

### Example

In this example we use the annual maxima of the daily maximal
temperature (TX) in Dijon (France) provided as the `TXMax_Dijon` data
frame


```{r, DijonDat, message=FALSE, warning=FALSE}
library(NSGEV)
head(TXMax_Dijon)
```
Note that the warnings and messages will not be shown in this example.

A TVGEV model requires a date variable beginnings of the annuela blocks 


```{r Dijon0, message=FALSE, warning=FALSE}
df <- within(TXMax_Dijon, Date <- as.Date(paste0(Year, "-01-01")))
fit0 <- TVGEV(data = df, response = "TXMax", date = "Date",
              loc = ~ 1)
coef(fit0)
autoplot(fit0)
```

We can fit a model with a linear time trend
	
```{r, Dijon1, message=FALSE, warning=FALSE}
fit1 <- TVGEV(data = df, response = "TXMax", date = "Date",
              design = polynomX(date = Date, degree = 1),
              loc = ~ t1)
autoplot(fit1)
coef(fit1)
anova(fit0, fit1)

```

The `predict` method can be used to compute conditional return levels
corresponding to a given year, be it a past or future year

```{r, DijonPred}
pred <- predict(fit1)
autoplot(pred)
autoplot(predict(fit1, confint = "proflik", trace = 0))
autoplot(predict(fit1, newdate = "2040-01-01", confint = "proflik", trace = 0))
```

The default confidence intervals are obtained by using the "delta
method" but profile likelihood intervals can be otained as well.


Rather than considering the maximum on a specific block, one can
consider the maximum $M$ on a larger period as sometimes called a *design
life period*. The distribution of $M$ is then no longer a GEV 


```{r, DijonMax, warning=FALSE, message=FALSE}
date <- as.Date(paste0(2025:2040, "-01-01"))
qm <- quantMax(fit1, date = date, confint = "proflik", trace = 0)
autoplot(qm) + ggtitle("Quantile of the maximum on 2025-2040")
```

By changing the value of the `design` argument one can different use
basis functions such as splines with given knots.


## INSTALLATION 

### With the *remotes* package 

In an R session use

```{r, eval=FALSE}
library(remotes)
install_github("IRSN/NSGEV", dependencies = TRUE)
```

This should install the package and make it ready to use.

Mind that by default this does not build the vignettes shipped with
the package (long-form documentation). To build the vignettes, use
instead

```{r, eval=FALSE}
install_github("IRSN/NSGEV", dependencies = TRUE, build_vignettes = TRUE)
```
The installation will then take more time but the vignettes will be
accessible from the help of the package (link above the "Help Pages"
section).

You can also select a specific branch or a specific commit by using
the suitable syntax for `install_github`. See the **remotes** package
documentation for more details.

### Precompiled versions

Precompiled versions of the package are available for some platforms
and can be downloaded by using the **Releases** link of the GitHub
Repository.  Use `NSGEV_x.y.z.tgz` for MacOS, `NSGEV_x.y.z.tgz` for
Windows and `NSGEV_x.y.z_R_x86_64-pc-linux-gnu.tar.gz` for Linux
Ubuntu where `x`, `y` and `z` are the major, minor and patch numbers
of the package version.
