---
title: "TVGEV: Check Return Levels"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{TVGEV: Check Return Levels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
    cache = FALSE,
    fig.width = 6, fig.height = 4, fig.path = "Rgraphics/")
KABLE <- TRUE
ip <- installed.packages()
library(NSGEV)
library(ismev)
```

## The check function

The Non-Stationary Return Level (NSRL) of Parey et al is implemented
for Time-Varying GEV models in the `predictUncond` function of the
**NSGEV** package. The package has a function named
`check_predictUncond` that can be used to check the results of
`NSGEV::predictUncond`. For now, the check does not check the
computation of the confidence limits, but only the value of the RL as
given by the `Quant` column in the data frame returned by
`predictUncond`. Since the NSRL is given by an equation with no
explicit solution, it is worth checking that the returned value is
consistent with the definition of the NSRL.

The check relies on the `simulate` method for the `"TVGEV"` class.
Firstly, an unconditional prediction is obtained for each period in
a set of periods of interest. All these periods share the same
beginning date, say $t_0$. The period with number $j$ contains $B_j$
blocks where $B_j$ is the maximal horizon chosen via `period` (or set
by default if `period` is not given). Remind that `predictUncond`
generally computes the results for *several* periods with the same
origin $t_0$ as required here. Secondly, a large number, say $m$, of
paths $Y^{[i]}(t)$ is generated from the model. For each predicted
period $j$ and each path $i$, we compute the number $N^{[i]}(u_j)$ of
exceedances over the quantile $u_j$ corresponding to the period. We
eventually take the average of the $m$ values $N^{[i]}(u_j)$ for $i=
1$ to $m$ and compare it to the expectation. If $m$ is large, then the
average should be close to the expectation.


## Example: annual maxima of TX in Dijon 
    
The *example* run with `example(TVGEV)` defines a `TVGEV` model for
the annual maxima of daily maximum temperature (TX) at Dijon (Fr) in
an object named `res2`.

```{r predictUncond, results = "hide", echo=FALSE}
example(TVGEV)
```
Consider as period of interest the period starting at year 2020. The
following code computes the NSRL for several periods of time.
	 
```{r predictUncond1, results = "asis", fig.cap="Unconditional Return Levels with 95% confidence intervals (profile-likelihood)."}
pu <- predictUncond(res2, newdateFrom = "2020-01-01", confintMethod = "proflik")
kable(pu, digits = 2)
plot(pu)
```
	
For instance, for the period of $B_j = 40$ years beginning at the date `"2020-01-01"`,
the level  $u_j =`r round(subset(pu, Period == 40)$Quant, dig = 2)`$ Celsius
is expected to be exceeded once (in average) across a large number of paths.
   
```{r predictUncond2}
set.seed(12345)
check_predictUncond(res2, newdateFrom = "2020-01-01")
```	
We use the default number of paths $m = `r formals(NSGEV::check_predictUncond)$nsim`$, but
only the `100` first paths are displayed on the plot. The results are in good
accordance with the definition.

## Findings

This test shows that the NSRL computed by `predictUncond` is in good
accordance with the definition.
	  
