---
title: "TVGEV: Check Profile Likelihood"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{TVGEV: Check Profile Likelihood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
    cache = FALSE,
    fig.width = 6, fig.height = 4, fig.path = "Rgraphics/")
KABLE <- TRUE
ip <- installed.packages()
library(NSGEV)
library(ismev)
```

## Goal
    
In this report, we compare profile-likelihood inference results given
by **NSGEV** with those of other packages, mainly **ismev**. We make
use of classical data.

## Port Pirie data

### ML Estimation
   
``` {r proflik-pp1}
library(ismev)   
data(portpirie, package = "ismev")
df <- portpirie
df <- within(df, date <- as.Date(sprintf("%4d-01-01", Year)))
fitNSGEV <- try(TVGEV(data = df, response = "SeaLevel", date = "date"))
fitismev <- gev.fit(xdat = df$SeaLev)
fitextRemes <- extRemes::fevd(x = df$SeaLev, type = "GEV")

rbind("NVSGEV" = coef(fitNSGEV),
      "ismev" = fitismev$mle,
      "extRemes" = fitextRemes$results$par)

```
	
### Confidence Intervals on Parameters

The `check_confint` function checks the computation of the
profile-likelihood intervals on the shape parameter. This is a
graphical test: on the plot produced by `ismev::prof.xi`, we add two
green vertical lines showing the confidence limits for the shape
parameter given by the `confint` method for the class `"TVGEV"`. We
must check that both lines cut the profile-likelihood curve at a point
with an ordinate matching the lowest horizontal blue line.
	    	
``` {r proflik-pp2a, fig.width=4, fig.height=4, fig.cap="Confidence level 95%."}    
check_confint(fitNSGEV, ref = fitismev, level = 0.95)
```
	    	
``` {r proflik-ppb, fig.height=4, fig.cap="Confidence level 70%."}    
check_confint(fitNSGEV, ref = fitismev, level = 0.70)
```
	
### Confidence Intervals on Return levels
    
The `check_predict` function checks the computation of the
profile-likelihood confidence intervals on return levels. As for
`check_confint`, the check is essentially graphical. On each plot, the
two vertical lines show the confidence limits given by the `predict`
method for the class `"TVGEV"`. We must check that both lines cut the
profile-likelihood curve at a point with an ordinate matching the
lowest horizontal blue line.
       
``` {r proflik-pp3a, results="hide", fig.width=8, fig.height=8, fig.cap="Confidence level 95%."}    
check_predict(fitNSGEV, ref = fitismev, level = 0.95)

```
		
``` {r proflik-pp3b, results="hide", fig.width=8, fig.height=8, fig.cap="Confidence level 70%."}    
check_predict(fitNSGEV, ref = fitismev, level = 0.70)

```

Note that the function is rather verbose, so the results have been hidden.

## Fremantle data

### ML Estimation
   
``` {r proflik-fmt1}
library(ismev)   
data(fremantle, package = "ismev")
df <- fremantle
df <- within(df, date <- as.Date(sprintf("%4d-01-01", Year)))
fitNSGEV <- try(TVGEV(data = df, response = "SeaLevel", date = "date"))
fitismev <- gev.fit(xdat = df$SeaLev)
fitextRemes <- extRemes::fevd(x = df$SeaLev, type = "GEV")

rbind("NVSGEV" = coef(fitNSGEV),
      "ismev" = fitismev$mle,
      "extRemes" = fitextRemes$results$par)

```
	
### Confidence Intervals on Parameters

	    	
``` {r proflik-fmt2a, fig.cap="Confidence level 95%.", fig.height=4, fig.width=4}    
check_confint(fitNSGEV, ref = fitismev, level = 0.95)
```
	
``` {r proflik-fmt2b, fig.cap="Confidence level 70%.", fig.width=4}    
check_confint(fitNSGEV, ref = fitismev, level = 0.70)
```
	
### Confidence Intervals on Return levels
		
``` {r proflik-fmt3a, results="hide", fig.width=8, fig.height=8, fig.cap="Condidence level 95%."}    
check_predict(fitNSGEV, ref = fitismev, level = 0.95)

```
	
``` {r proflik-fmt3b, results="hide", fig.width=8, fig.height=8, fig.cap="Confidence level 70%."}    
check_predict(fitNSGEV, ref = fitismev, level = 0.70)

```
	
## Findings

It can happen that one of the bound of a confidence interval is not
found and remains `NA`. This seems to be most often the case for the
lower bound.	  
