
---
title: "Using variables in the `design` argument of `TVGEV`"
author: "Yves Deville <deville.yves@alpestat.com> and Nathan Bengaouer <nathan.bengaouer@asnr.fr>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
linkcolor: blue
vignette: >
  %\VignetteIndexEntry{Using variables in the dscign argument of NSGEV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\newcommand{\m}[1]{\mathbf{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}

# Using variables in the `design` argument

The `design` argument of the `TVGEV` function is intended to be
"language". It can embed a "parameter" given as a R object in
the environment, yet this can cause problems due to the scoping
rules, especially when the `TVGEV` function is used from within a
function.

# Example change of slope in a broken line trend

As an example, we will consider the case where a time trend is
specified as a broken line with a slope zero before the time of
change, say $u$, and with an unknown slope after the time $u$. This is
similar to a *kink regression* in the linear regression framework.

It is not possible with `{NSGEV}` to consider $u$ as a model parameter
because `TVGEV` models are assumed to link each of the three GEV
parameters to a linear predictor having the form $\m{x}(t)^\top
\bs{\psi}$. Incidentally, would $u$ be considered as a model
parameter, the log-likelihood would no longer be differentiable
w.r.t. $u$, and adjustments would be required in the likelihood-based
inference. However we can think of using a grid of candidate values
for the change of slope time $u$: by fitting the model with $u$ fixed
to each grid value say $u_i$ for $i=1$, $2$, ... we can compute the
corresponding maximised log-likelihoods and then choose the value of
$u$ as the grid value with maximal log-likelihood. It is convenient to
write a function inside which the loop on the candidate values is
found. The broken line can be obtained with the `breaksX` design
function: its `breaks` argument corresponds to the vector of knots for
the truncated power splines. By using a single knot, one gets two basis
functions - see the help for `breaksX`. One only needs to use the
broken line basis function with name `"t_1990"` if for example the
change of slope $u$ is located at year 1990. A possible problem arises
from the fact that the candidate value $u_i$ has to be passed to the
design function.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(NSGEV)
```

For illustration, consider the `Dijon_TXmax` example: assume that
the plausible breaks are in the range `1970` to `2000`.
To make things slightly more general we assume that `col` contains
the name of the column for the maxima in the data frame `maxData`.


```{r fig.width=8, fig.height=6, fig.align="center", out.width="70%"}
fitBreaks <- function(maxData, col) {   
    if (!inherits(maxData$year, "Date")) {
        maxData$year <- as.Date(paste0(maxData$year, "-01-01"))
    }
    res <- desCall <- list()
    yearBreaks <- c(1970:2000)
    for (ib in seq_along(yearBreaks)) {
        if (TRUE) { ## WORKS 
            desCall[[ib]] <-
                sprintf("breaksX(date = year, breaks = \"%4d-01-01\", degree = 1)",
                        yearBreaks[ib])
            floc <- sprintf("~ t1_%4d", yearBreaks[ib])
            text <-
                sprintf(paste0("TVGEV(data = maxData, response = col, date = \"year\",\n",
                               "       design = %s,\n",
                               "      loc = %s)\n"),
                        desCall[[ib]], floc)
            res[[ib]] <- eval(parse(text = text))
        } else { ## DOES NOT WORK
            d <- sprintf("%4d-01-01", yearBreaks[ib])
            floc <- sprintf("~ t1_%4d", yearBreaks[ib])
            res[[ib]] <- TVGEV(data = maxData, response = col, date = "year",
                               design = breaksX(date = year, breaks = d, degree = 1),
                               loc = floc)
        }
    }
    ll <- sapply(res, logLik)
    iMax <- which.max(ll)
    ## Compute lims to include the 'confidence' logLik level
    llLim <- ll[iMax] - qchisq(0.95, df = 1) / 2
    yearMax <- yearBreaks[iMax]
    plot(yearBreaks, ll, type = "o", pch = 21, col = "orangered",
         ylim = range(ll, llLim)                               ,
         lwd = 2, bg = "gold", xlab = "break", ylab = "log-lik")
    grid()
    iMax <- which.max(ll)
    abline(v = yearMax)
    abline(h = ll[iMax] - c(0, qchisq(0.95, df = 1) /2),
           col = "SpringGreen3", lwd = 2)
    fit <- res[[iMax]]
    return(fit)  
}

names(TXMax_Dijon) <- c("year", "TXMax") 
(fitL <- fitBreaks(maxData = TXMax_Dijon, col = "TXMax"))
```

# Explanation

The object given in the `design` formal argument is normally a *call*
to a function, which is internally used with `do.call`. If a R object
such as `d` is used in the call, it is first looked for in the
environment of the call (here, the data frame), then in the
environment of the function (here, `fitbreaks`). In both cases the
object `d` does not exist.

We could turn `d` into a (constant) column of the data fame: this
would make the estimation successful. Yet the fitted object would
require for a prediction with `predict` that the column `d` is
found in the provided `new` data, which is tedious. The solution
chosen here is to parse a code with the variables replaced by
their value. We see that the returned `TVGEV` object contains the
wanted formula and the wanted design. It does no longer relate to
an extra R variable.

From a statistical perspective, it must be kept in mind that the
ML estimate of the break does not have the standard normal
distribution, because the log-likelihood function is not
differentiable. On intuitive grounds, the estimate should be
"super-efficient". Using the classical chi-square approximation to
derive a profile likelihood interval should consequently lead to
an interval with a coverage rate that is larger than the nominal
one. On the plot the horizontal green lines show the maximised
log-likelihood $\ell_{\texttt{max}}$ and the value $\ell_{\texttt{max}} -
\delta$ that would be used to find the limits of a profile
likelihood confidence interval under the classical regularity
conditions, which are not met here.


