---
title: "TVGEV: Check Venice Example"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{TVGEV: Check Venice Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
    cache = FALSE,
    fig.width = 6, fig.height = 4, fig.path = "Rgraphics/")
KABLE <- TRUE
ip <- installed.packages()
library(NSGEV)
library(ismev)
```
	
### Venice example and data

This example was studied by Coles and
[Fawcett](http://www.mas.ncl.ac.uk/~nlf8/teaching/mas8391/background/chapter5.pdf)
who both fitted time varying GEV models with linear and quadratic
trend by using the **ismev** package. We reproduce here their results
by using **NSGEV** and also **extRemes**.

Although the data contains the $r$-largest observations, we use only
the annual maxima here; These are found in column 2 of the matrix
object.
      
     
``` {r venice1}    
data(venice, package = "ismev")
df <- data.frame(Year = venice[ , 1],
                 date = as.Date(sprintf( "%4d-01-01", venice[ , 1])),
                 SeaLevel = venice[ , 2])
df <- within(df, date <- as.Date(sprintf("%4d-01-01", Year)))
plot(SeaLevel ~ date, data = df, type = "o", pch = 16, xlab = "")

```
	
### Fitting stationary GEV models    

We begin by fitting a stationary GEV model.
    
``` {r veniceNoTrend, results="hide"} 
estim <- "optim"
fitTVGEV0 <- TVGEV(data = df,
                   response = "SeaLevel", date = "date",
                   estim = estim,
                   loc = ~ 1)
fitismev0 <- gev.fit(xdat = df$SeaLev)
fitextRemes0 <- extRemes::fevd(x = df$SeaLev, type = "GEV")
```	

The estimated coefficients and maximised log-likelihoods are as follows.
	
``` {r veniceNoTrend1} 
rbind("NSGEV" = coef(fitTVGEV0),
      "ismev" = fitismev0$mle,
      "extRemes" = fitextRemes0$results$par)
c("NSGEV" = logLik(fitTVGEV0),
  "ismev" = -fitismev0$nllh,
  "extRemes" = -fitextRemes0$results$value)


```

### Linear trend

We now introduce a linear trend. We first create a dataframe to be
used with the functions from **extRemes** and **ismev**, with a linear
and variable a quadratic trend variable as well.
	
``` {r veniceLinTrend, results="hide"} 
dfismev <- within(df, {t1num <- Year - 1930; t2num <- (Year - 1930)^2 })
fitTVGEV1 <- TVGEV(data = df,
                   response = "SeaLevel", date = "date",
                   design = polynomX(date = date,
                       degree = 2, origin = "1930-01-01"),
                   estim = estim,
                   trace = 2,
                   loc = ~ 1 + t1)
fitismev1 <- gev.fit(xdat = df$SeaLev,
                     ydat = as.matrix(dfismev[ , "t1num"]),
                     mul = 1)
fitextRemes1 <- extRemes::fevd(x = dfismev$SeaLev,
                               data = dfismev,
                               type = "GEV",
                               location.fun = ~t1num)
```	

``` {r veniceLinTrend1} 
rbind("NSGEV" = coef(fitTVGEV1),
      "ismev" = fitismev1$mle,
      "extRemes" = fitextRemes1$results$par)
c("NSGEV" = logLik(fitTVGEV1),
  "ismev" = -fitismev1$nllh,
  "extRemes" = -fitextRemes1$results$value)

```
We can perform a likelihood-ratio test as in the references.
   	
``` {r veniceLinTrend2} 
anova(fitTVGEV0, fitTVGEV1)

```

### Quadratic trend

We now add a quadratic term in the trend.    
    
``` {r veniceQuadTrend, results="hide"} 
fitTVGEV2 <- TVGEV(data = df,
                   response = "SeaLevel", date = "date",
                   design = polynomX(date = date,
                       degree = 2, origin = "1930-01-01"),
                   estim = estim,
                   trace = 2,
                   loc = ~ 1 + t1 + t2)
fitismev2 <- gev.fit(xdat = dfismev$SeaLev,
                     ydat = as.matrix(dfismev[ , c("t1num", "t2num")]),
                     mul = c(1, 2))
fitextRemes2 <- extRemes::fevd(x = dfismev$SeaLev,
                               data = dfismev,
                               type = "GEV",
                               location.fun = ~t1num + t2num) 
```	
The estimated coefficients and maximised log-likelihoods are as follows.

``` {r veniceQuadTrend1} 
rbind("NSGEV" = coef(fitTVGEV2),
      "ismev" = fitismev2$mle,
      "extRemes" = fitextRemes2$results$par)
c("NSGEV" = logLik(fitTVGEV2),
  "ismev" = -fitismev2$nllh,
   "extRemes" = -fitextRemes2$results$value)

```
So, compared to the two other packages, `NSGEV::TVGEV` returns different estimated coefficients
and a smaller maximised log-likelihood, indicating a problem in the maximisation. Since
the scaling of the trend variable is often a concern, we try a change in the time origin.
	
``` {r veniceQuadTrend2} 	
fitTVGEV2b <- TVGEV(data = df,
                    response = "SeaLevel", date = "date",
                    design = polynomX(date = date,
                        degree = 2, origin = "1950-01-01"),
                    estim = estim,
                    trace = 2,
                    loc = ~ 1 + t1 + t2)
rbind("NSGEV" = coef(fitTVGEV2b),
      "ismev" = fitismev2$mle,
      "extRemes" = fitextRemes2$results$par)
c("NSGEV" = logLik(fitTVGEV2b),
  "ismev" = -fitismev2$nllh,
  "extRemes" = -fitextRemes2$results$value)
```
The log-likelihood maximised with `NSGEV::TVGEV` is now identical to that
of the other two packages so a simple change of origin solved the
problem. The resulting model in the `TVGEV` object is simply a
re-parameterisation of the former, and the comparison with the other
packages needs some care because the trends for the location parameter
are no longer identical. The new parameterisation facilitates the
maximisation. To complete the comparison we can use 1950 as an origin
for all models.
    
``` {r veniceQuadTrend3, results="hide"} 
dfismevb <- within(df, {t1num <- Year - 1950; t2num <- (Year - 1950)^2 })
fitismev2b <- gev.fit(xdat = dfismevb$SeaLev,
                      ydat = as.matrix(dfismevb[ , c("t1num", "t2num")]),
                      mul = c(1, 2))
fitextRemes2b <- extRemes::fevd(x = dfismevb$SeaLev,
                                data = dfismevb,
                                type = "GEV",
                                location.fun = ~t1num + t2num)
```
``` {r veniceQuadTrend4} 
rbind("NSGEV" = coef(fitTVGEV2b),
      "ismev" = fitismev2b$mle,
      "extRemes" = fitextRemes2b$results$par)
c("NSGEV" = logLik(fitTVGEV2b),
  "ismev" = -fitismev2b$nllh,
  "extRemes" = -fitextRemes2b$results$value)

```
We now can proceed to the likelihood ratio test.
``` {r veniceLinTrend4} 
anova(fitTVGEV1, fitTVGEV2b)

```
	
The test tells that adding a quadratic trend term does not improve the
model.

	
### Findings
	
With this example it appears that in `NSGEV::TVGEV` the maximisation of
log-likelihood can fail to find the best value, still giving an
indication of convergence. Here the problem is due to the order of
magnitude of one covariate, namely the quadratic term in the
polynomial basis. It was fixed by simply changing the origin of time.

With the current implementation of **NSGEV** when formulas involving
several basis covariates are used, we recommend to fit models with
increasing complexity, and check that the log-likelihood increases
when adding a new covariate.

Several solutions can be to improve the code of `NSGEV::TVGEV` in
future versions.

- Allow the user to give initial values for the parameters.

- Try a "multistart" optimisation with randomly selected initial
  values.
		
- Scale the design matrix $\mathbf{X}$ by subtracting its mean to
  each column, and by dividing each column by its standard
  deviation. Although this is a quite simple application of the
  `scale` method for the `"matrix"`class,  the estimated coefficients
  should then be translated back in the original scale, which can be
  quite tedious to implement.

  
